---
output:
  pdf_document: 
    citation_package: natbib
    fig_caption: yes
    template: ./resources/aog-latex-ms.tex
  word_document:
    reference_docx: ./resources/aog_word_style.docx
fontfamily: mathpazo
fontsize: 11pt
anonymous: false
endnotes: false
pagenumber: 1
bibliography: ../project.bib
biblio-style: ./resources/ajs.bst
title: "Does the Media Influence Voters: A Study of the 2012 and 2016 Elections"
author:    
- affiliation: University of Oregon, Political Science
  name: Mary Follo 
- affiliation: University of Oregon, Sociology 
  name: Aaron Gullickson
keywords: Media Influence, Voting, Elections
thanks: Thank you to the University of Oregon, Political Science department for forcing me to take this class and the Soc 513 family for showing up every Tuesday and Thursday with sub-par enthusiasm. Also a big thanks to the UO Econ department who showed me a million different ways to work with R, all which confused me more. Shoutout to Chris from Econ for being GitHub friends with me. And a final huge thank you to Aaron who put up with so much, you're the real MVP. 
abstract: As the Unites States has become more polarized with intense and contested elections, understanding the impact that variables like media have on voting behavior are becoming more and more important. This study uses ANES data from the 2012 and 2016 election to compare the effect that attention to news media as well as age and party affiliation has on voting. I found small effects that lack of media attention does result in higher rates of voting for a republican candidate.  Additionally, data shows that younger Democrats are more likely to follow news media than older Democrats and older Republicans are more likely to pay attention than younger Republicans. Further I was able to find that media attention has increased between the two elections, proving that greater research is needed to determine how elections can be influenced by exogenous variables like media. 
---

# Introduction

The role of media in voting has always played a significant role in U.S. politics. From political propaganda pamphlets during the revolutionary war to the political machine era; parties, candidates and their campaigns have worked to exert influence over voters. Today, the media is one of the strongest resources candidates have at their disposal to mobilize their voters. All aspects of campaigns are broadcast on television, the Internet and social media in an attempt to influence behaviors. In fact, recent research, when studying the effect of Fox news bias, has shown that the media can have a sizable political impact (DellaVigna and Kaplan, 2006). In the 2008 election for example, Oprah Winfrey’s celebrity endorsement of Barack Obama on her television show can be correlated with an extra one million votes during the election. Hence, the “Oprah Effect” or ability for democracy to be strengthened among low-awareness voters through the consumption of soft news media was created (Baum, 2006). At the time of the 2008 election however, a new media type was just taking shape. Social media that is now heavily utilized for campaigns was just becoming accessible to users with Facebook and Twitter taking shape in 2006 and Instagram in 2008. These new forms of social media are just one more way that candidates and campaigns can exert influence over voters. 

Understanding the role of media in public opinion and behavior has historically been hotly debated among three distinct models: the hypodermic model, the minimum effects model and the constructivist model. The hypodermic model asserts that “the mass media exercised a powerful and persuasive influence” on a public that was “inherently susceptible to manipulation” (Curran, Gurevitch and Woollacott, 1982). Post-World War I, scholars theorized that the media was seen as a brainwashing agent that preyed on the unknowing public mind that could be completely consumed by media agency.  This pessimistic theory has proven empty over time and does not take into account individual agency but instead assumes each person is a blank slate, a tabula rasa that is waiting for the media or some other influence to give them opinions. In the 1950’s and 60’s a new school of thought emerged called the minimal effects theory. Introduced by Joseph Klapper, the minimal effects theory finds that “only a tiny fraction of voters actually changed their vote intentions during an election campaign, that audience motivations and prior beliefs influenced the interpretations of persuasive messages” (Neuman& Guggenheim, 2011). The idea of media manipulation shifted from “what the media does to people to what the people do with the media” (Gamson, 1988). In this way, the minimal effects approach tells a narrative about a voter with more agency but does not take into account the real role that media framing and priming play in political behavior. These two early theories of media influence portrayed the American citizen as an all of nothing individual, either completely consumed by the media or completely obstinate to it. 

Instead, I focus on the use of the constructivist model that “takes into account the changing nature of the media’s power as it interacts among voters who influence each other while contributing to the construction of messages, meanings and outcomes in the electoral process” (Gamson, 1988). The constructivist model allows one’s personal social discourse to align with the media discourse that one is influenced by to construct their own opinions and vote choice by melding information obtained from the media with their existing beliefs (Neuman et. al, 1992). In this way, citizens are able to consume a large variety of news media and construct their own political knowledge out of what they believe is most relevant. The agenda setting and priming that is absent in the minimal effects approach is welcomed in the constructivist approach and shows that the media can be both directly and indirectly persuasive through the frames it chooses to shape discourse with (Armoudian & Crigler, 2010). As technology continues to develop, more factors will be added to the complex interplay that constructs each individual’s political discourse. Political meaning and the construction of personal voting behavior then is not solely determined by media effect, nor does it have a minimal effect. Media, especially social media can be effective helping shape one’s thoughts through its news generating ability, however when it comes to actual decision making and voter behavior, individuals tend to rely on their interpersonal networks for direction. 

As mass media like social media have gained prominence in American society, citizens have become dependent upon them for a variety of news information. The reliance on social media especially by youths for news relates to Ball-Rokeach and DeFleur’s dependency theory. Dependency theory states that, “audiences are dependent on media information resources which leads to modifications in both personal and social processes” (Ball-Rokeach and DeFleur, 1976). What I believe to be an extension of the constructivist model, dependency theory not only states media’s role in helping create discourses but it gives power to mass media by acknowledging the increasing reliance upon it for information and resources. “In American society, the media are presumed to have several unique functions. They operate as a Fourth estate gathering and delivering information about the actions of the government; […] they constitute the principle source of the ordinary citizen’s conceptions of national and world events; they provide enormous amounts of entertainment” (B-R and DeFleur, 1976). Gaining a wide variety of important information however does not mean that people are influenced, “Topics are filtered through media information-gathering and –processing systems and then selectively disseminated. The public then sorts out their interest and concern with this information as a function of both their individual differences in personal make-up and their location in societal strata and categories” (B-R and DeFleur, 1976). 

Despite what has been characterized in the literature by many American behavioralists, interpersonal effects influence a person’s decision-making and attitude more than the information they receive from the media. “American voters do not operate in the social vacuum that much of the contemporary voting literature seems to assume. Rather, voters’ enduring personal characteristics interact with messages they are receiving from the social context in which they operate […] interpersonal discussion outweighs the media in affecting the vote” (Beck et. al, 2002). Additionally, recent literature has focused more on the effect that position in one’s social network has a greater effect on voting behavior than does media influence.
“The dependencies people have on media information are a product of the nature of the socio-cultural system, category membership, individual needs, and then number and centrality of the unique information functions that the media system serves for individual and society […] when people’s social realities are entirely adequate before and during message reception, media messages may have little to no alteration effects” (B-R, DeFleur, 1976). 

Operating under a constructivist model and dependency theory one can begin to understand how media is just one factor of a greater social system that can shape one’s attitude and behavior. The 2012 and 2016 elections are two great case studies for this project because of the prominence of both traditional forms of media and social media. Given the attention paid to the 2016 election in regards to “fake news”, media bombshells and negative ads, we should expect media attention to have increased from the 2012 election. Within the scope of this paper, I believe that while media attentiveness has increased from 2012 to 2016, I do not believe this impacts one type of voter or one party more than the other. Further I believe that the accessibility of media to younger voters will result in more media attention paid by younger voters and expect media attention to taper off with older voters. 



# Data and Methods

For this project I used ANES Time Series data from both the 2012 and 2016 election surveys. The 2012 Time Series study data was collected between September 2012 and January 2013, with respondents being interviewed two months prior to the 2012 election as well as the day after the election. A voter validation data set was added and will be merged with the Time Series data to to validate self-reported voter turnout and registration from the Time Series data. This data includes content on electoral participation, voting behavior and public opinion. 

The 2016 Time Series study is similar to the 2012 but has additional features including supplemental data pulled from respondents Facebook accounts. Additionally, the Methodology file includes data on households, interviews and records of contact among respondents. The Address data gives voter turnout status for adults at each address that was sampled in the Time Series study. 

I created a subset of the following variables in the ANES 2016 dataset: vote choice, how often they use media news to follow politics, gender, age, state, party affiliation, how many days do they use social media and how closely they follow politics in the media. I created a subset of the following variavles in the ANES 2012 dataset: vote choice, state, gender, age, how often they use TV news to follow politics, how often they use the Internet to follow politics, party affiliation and how often they are interested in politics. 

After created subsets of each data set I recoded all the variables, in some cases I combined answers and omitted non-responses including turning age into a continuous variable. For my models I was able to subset the data again to include only voting ages and worked on combining rows and columns in certain variables to merge matrices. The graphs created include bivariate analysis of candidate choice across age groups as well as media attention across age groups and media attention by party affiliation. Finally, I used an interaction term to create a multinomial model to determine if voting by different levels of media attention was significant. 

# Results

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, fig.align = "center")
load("~/Desktop/soc513_spring_project/analysis/output/analytical_data.RData")
source("~/Desktop/soc513_spring_project/useful_functions.R")
library(ggplot2)
library(knitr)
library(haven)
library(readr)
library(texreg)
library(nnet)

```

```{r vote_age_2012, echo=FALSE, fig.cap="Graph 1: Proportion of Candidate Choice across Age Group Means 2012"}  
  ggplot(cleaned_data2012, aes(x=vote, y=..prop.., group=1))+
  geom_bar()+
  facet_wrap(~age)+
  coord_flip()+
  labs(x="Candidate Choice", y="Proportion")+
  theme_bw()
  
```

```{r vote_age_2016, echo=FALSE, fig.cap="Graph 2: Proportion of Candidate Choice across Age Group Means"}
ggplot(cleaned_data2016, aes(x=vote, y=..prop.., group=1))+
  geom_bar()+
  facet_wrap(~age)+
  coord_flip()+
  labs(x="Candidate Choice", y="Proportion")+
  theme_bw()
#More media attention padi by older people who also overwhelmingly vote for trump
```

The first two graphs show a bivariate analysis of the proportion of candidate choice across age group means. For this analysis, each age group was subset to include a range of five years. The 2012 graph shows that for all age groups, except age 72, Obama was the overwhelming favorite. The youngest group of voters chose to vote for Obama at much higher proportions (40% compared to 10%). While the proportion of votes for Romney begins to increase when a person enters their 40’s, Obama is there clear winner in almost all age groups. The 72-year-old age group does show a slightly higher probability of voting for Romney compared to Obama, but for the 80 year old age group; there are equal proportions of votes for each candidate. In the 2016 data, proportions are closer between the two candidates with Clinton capturing around 25% of votes from the youngest age group while Trump received around 15%. The proportion of votes for Trump increases at an early age than did votes for Romney in 2012. In the age range of 47, one becomes slightly more likely to vote for Trump. While proportions fluctuate, Trump maintains the advantage for most age groups over 42. This of course is no surprise for most scholars of American voting behavior. The Republican Party overwhelmingly targets an older demographic with more conservation policies than the Democratic party. What is interesting however is that the proportion difference is much closer in 2016 than it was in 2012, this can be for a variety of reasons including incumbent advantage or the progressive pull of Obama’s neo-liberal policies on Republicans. What it does appear to confirm however is that the older one gets, the more likely one is to vote Republican. 

```{r media_age_2012, echo=FALSE, fig.cap= "Graph 3: Proportion of Media Attention for Internet Users Across Age Group Means 2012"}
ggplot(cleaned_data2012, aes(x=media_intINT, y=..prop.., group=1))+
  geom_bar()+
  facet_wrap(~age)+
  coord_flip()+
  labs(x="Media attention by Internet Users", 
       y="Proportion")+
  theme_bw()
```

```{r media_age_2016, echo=FALSE, fig.cap= "Graph 4: Proportion of Media attention across Age Group Means 2016"}

ggplot(cleaned_data2016, aes(x=media_att, y=..prop.., group=1))+
  geom_bar()+
  facet_wrap(~age)+
  coord_flip()+
  labs(x="Media attention", y="Proportion")+
  theme_bw()
```
The next set of graphs explores the relationship between age and the amount of attention one pays to politics in the media. In 2012, for voters between the ages of 19 and 35, most paid little to no attention to politics in the media. As age increases, the amount of voters paying little to no attention decreases, as do most of the categories. This makes sense, as older people are less likely to use media like the Internet to consume information about politics. Further, in 2012, media was less accessible to older populations than it is now. In 2016, proportions for media attention are much higher overall. Still, little to none, and a moderate amount are the categories that best characterize young people and politics in the media. However, starting in one’s late 30’s paying attention to politics in the media increases and the 72 year old age group reporting the highest proportion of users paying “A lot” of attention. While this is a somewhat unusual result, I believe this can be attributed to media becoming the primary way people receive news. Out are of the days of newspaper, radio and even television news, as more and more people consumer Internet news. As media has become more accessible to older generations, it should not be surprising that older people use media to become informed about politics than younger generations who are more interested in popular apps and websites. 

```{r media_party_2012, echo=FALSE, fig.cap="Graph 5: Proportion of Media Attention by Internet Users across Parties 2012"}
ggplot(cleaned_data2012, aes(x=media_intINT, y=..prop.., group=1))+
  geom_bar()+
  facet_wrap(~party)+
  labs(x="Media Attention for Internet Users by Party", y="Proportion")+
  coord_flip()+
  theme_bw()
```

```{r media_party_2016, echo=FALSE, fig.cap= "Graph 6: Proportion of Media Attention across Parties 2016"}
ggplot(cleaned_data2016, aes(x=media_att, y=..prop.., group=1))+
  geom_bar()+
  facet_wrap(~partyreg)+
  coord_flip()+
  labs(x="Media Attention by Party", y="Proportion")+
  theme_bw()
```
My final bivariate analysis plots look at media attention by party affiliation. As shown above in Graphs 1and 2, voting Republican seems to increase over age and attention to politics using media also seems to increase by age. However, I also want to look at party affiliation and media interest. Do we see these same results by party and media attention, or are age, voting and media not as contingent on each other? Further, voting and media attention both have higher proportions in 2016 than in 2012, by exploring more variables, we can understand how much or how little of an impact the growing media has on voting. In 2012, Republicans paid greater attention to the media than did Democrats except in the categories of “A little” and “None”. This is interesting that Republicans would pay more attention gives that Obama won out in almost all ages. Furthermore, in 2016, Republicans have slightly higher rates of paying a great deal of attention, but Democrats pay more attention in the “A lot” and “A moderate amount” categories. Rates in 2016 are also much larger than rates in 2012 with each response gaining nearly 30% of responses compared to the 20% and under than answers received in 2012. What this could show is that the opposition party pays more attention to the news media than does the incumbent party. Further, in 2016, Democrats who were trying to mobilize a broad base, paid more attention than did homogenous Republicans. 

```{r model_2012, echo=FALSE, fig.cap="Graph 7: Proportion of Media Attention by Candidate Choice across Age"}

#
modelx.full_2012<-multinom(vote~media_intINT*I(age-40)+gender, data=cleaned_data2012)
round(coef(modelx.full_2012),4)

age <- 18:80
fakedata_12 <- data.frame(media_intINT=rep(c("A little/None","A moderate amount","A lot","A great deal"), length(age)),
                       age=rep(age,each=4), gender=rep("Female",4*length(age)))
predicted.probs_12 <- predict(modelx.full_2012, fakedata_12, type="probs")

fake.results_12 <- cbind(rbind(fakedata_12, fakedata_12, fakedata_12, fakedata_12),
                      as.data.frame.table(predicted.probs_12))
colnames(fake.results_12) <- c("media_intINT","age","gender","id","candidate","proportion")


ggplot(fake.results_12, aes(x=age, y=proportion, color=candidate, group=candidate))+
  geom_line()+
  facet_wrap(~media_intINT)+
  theme_bw()

ggplot(fake.results_12, aes(x=age, y=proportion, color=media_intINT, group=media_intINT))+
  geom_line()+
  facet_wrap(~candidate, scales= "free_y")+
  theme_bw()+
theme(legend.position="right")+
  labs(x="Age",
       caption="Proportion of Media Attention by Candidate Choice across Age", 
      col="Media Attention",
       y="Proportion")


```

```{r model, echo=FALSE, fig.cap="Graph 8: Proportion of Media Attention by Candidate Choice across Age"}
#create continuous age variable

model.full<-multinom(vote~media_att*(I(age-40))+gender, data=cleaned_data2016)
round(coef(model.full),5)

age <- 18:80
fakedata <- data.frame(media_att=factor(rep(c("A little/None","A moderate amount","A lot","A great deal"), length(age)),
                                        levels=c("A little/None","A moderate amount","A lot","A great deal")),
                       age=rep(age,each=4), gender=rep("Female",4*length(age)))
predicted.probs <- predict(model.full, fakedata, type="probs")

fake.results <- cbind(rbind(fakedata, fakedata, fakedata, fakedata),
                      as.data.frame.table(predicted.probs))
colnames(fake.results) <- c("media_att","age","gender","id","candidate","proportion")

ggplot(fake.results, aes(x=age, y=proportion, color=candidate, group=candidate))+
  geom_line()+
  facet_wrap(~media_att, scales = "free_y")+
  theme_bw()

ggplot(fake.results, aes(x=age, y=proportion, color=media_att, group=media_att))+
  geom_line()+
  facet_wrap(~candidate, scales = "free_y")+
  theme_bw()+
  theme(legend.position="right")+
  labs(x="Age",
       caption="Proportion of Media Attention by Candidate Choice across Age", 
      col="Media Attention",
       y="Proportion")

```
Graphs 7 and 8 show the proportion of media attention by candidate choice across age with the interaction effect of age and controlling for gender. With the interaction effect, we get a slightly different graph than shown above. With younger voters having a high proportion of media attention when voting for Obama. The highest media attention and the lowest converge at around 70 years old and begin to drop off. For Romney voters however, young people have lower proportions with the highest and lowest media attention both around 30 years of age. As the age of Romney voters increases, so does attention to news media. Further, those who pay a great deal of attention and those who pay no attention are not as divergent as the rates for Obama supporters. This shows that in general Romney voters pay less attention to news media than do Obama voters. In 2016, Clinton supporters like the Obama supporters in 2012, are more engaged in news media at younger ages. In fact, news media attention decreases where it begins to converge around the age of 80; much older than that of Obama supporters. Trump, like Romney, has the opposite effect with the highest rate of media attention being “A little to none” at nearly 40% for young voters. While a little to none maintains being one of the highest rates across age, all media attention increases over age, converging around 70 years old. What this shows is that for Trump supporters is that being uninformed characterizes many young voters and again that voting for a Republican increases as one ages. Unlike Romney voters whose rates of any media attention were relatively low for young voters, Trump supports paid more attention to news media at all ages. This could be indicative of increasing availability of media as well as the amount of media sources that now cover politics. 

```{r model_attempt, echo=FALSE, fig.cap="Table 1: Regression Model Predicting Voting by Attention Paid to the Media across Age"}

model.full<-multinom(media_att~partyreg+vote+age, data=cleaned_data2016)
coef(model.full)

screenreg((model.full), 
custom.coef.names=c("A great deal",
                    "A lot",
                    "A moderate amount", 
                    "Interaction (age-40)", 
                    "Male", 
                    "A great deal (age-40)", 
                    "A lot (age-40)", 
                    "A moderate amount (age-40)"), 
caption="Regression Model Predicting Voting by Attention Paid to the Media across Age", 
caption.above=TRUE)
 

```
Finally looking only at the 2016 election, Table 1 shows a regression of media attention by party registration holding vote choice and age constant. In this table, voting for Clinton and being female is the constant. What we see is that for “A great deal” and “A lot” of media attention, as those rates go down, voting for Trump increases. Put differently, paying less attention to news media makes one more likely to vote for Trump. These two variables are statistically significant. Additionally, being male and being older makes one more likely to vote for Trump. At the bottom of Table 1, paying “A moderate amount” of media attention with the interaction of age is shown to be significant, because this result includes all of the other variables in the model, it is likely that collinearity produced this result. 

# Conclusions

Hence, what the data tells us is that media effects matter based on age and party affiliation. Young democrats pay more attention to the media to get their political news while older democrats play less attention to forms of media. Alternatively, young republicans pay little attention to the media for political news while older republicans pay a great deal more attention. Possible reasons for these results include the fact that the Democratic Party targets younger people with more progressive policies and is able to better mobilize influential groups and people to help campaign. We see this mobilization of young people through celebrity endorsement and campaign related text communication. Republicans on the other hand promote far more conservative policies that appeal to an older crowd. Specifically in 2016, young people not paying attention to politics in the media helped garner Trump support in the polls, as he was the subject of many controversies. Young people who got their news from older relatives or like-minded friends would not receive the same political information as those who watched the news or even read twitter. Therefore, this data confirms the bias and stereotypes of the different parties when gathering information to vote. While this result is plausible, more data on voting behavior is needed to correlate the results. Further the lack of specific behavior based questions by the ANES could result in skewed data. Given the fact the “not available” or no answer was a widely answered response in the surveys, it is possible that this impacted the results especially in the data on proportions. With more time, a factor analysis would have been beneficial as well as using a multiple imputation model to impute the extensive missing values. Finally, more data is needed to understand the impact of specific types of media on voters. 

Therefore, this project confirmed my hypothesis that media impact has increased since 2012 and of media effects, but only for the Democratic Party. My hypothesis that media would not affect one party more than the other is essentially disproved by this data, as results were different between the two parties. More research is needed on the Republican Party and how their information is gathered specifically among younger voters. Looking back on the constructivist model of media effects, it is possible that younger voters gather political information from interpersonal sources as opposed to the media, which results in their voting behavior. As campaigns and parties become more polarized and contested and the role of media in our lives grow, understanding how different variables like how the media impacts political behavior will be key to American politics. 



# References

Armoudian, M., Crigler, A. N., & Leighley, J. E. (2010). Constructing the vote: Media effects in a constructionist model. In The Oxford handbook of American elections and political behavior. Oxford University Press.

Ball-Rokeach, S. J., & DeFleur, M. L. (1976). A dependency model of mass-media effects. Communication research, 3(1), 3-21.

Baum, M. A., & Jamison, A. S. (2006). The Oprah effect: How soft news helps inattentive citizens vote consistently. The Journal of Politics, 68(4), 946-959.

DellaVigna, S., & Kaplan, E. (2007). The Fox News effect: Media bias and voting. The Quarterly Journal of Economics, 122(3), 1187-1234.

Gamson, W. A. (1988). The 1987 distinguished lecture: A constructionist approach to mass media and public opinion. Symbolic interaction, 11(2), 161-174.

Garthwaite, C., & Moore, T. J. (2012). Can celebrity endorsements affect political outcomes? Evidence from the 2008 US democratic presidential primary. The journal of law, economics, & organization, 29(2), 355-384.

Gurevitch, M., Bennett, T., Curran, J., & Woollacott, J. (Eds.). (1982). Culture, society and the media (Vol. 759). London: Methuen.

Klapper, J. T. (1960). The effects of mass communication.

Neuman, W. R., & Guggenheim, L. (2011). The evolution of media effects theory: A six-stage model of cumulative research. Communication Theory, 21(2), 169-196.

Zaller, J. R. (1992). The nature and origins of mass opinion. Cambridge university press

