---
title: "Analysis for Project"
output: 
  html_document: 
    fig_height: 6
    fig_width: 9
    toc: yes
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message= FALSE, fig.align = "center")
source("useful_functions.R")
load("output/analytical_data.RData")
library(ggplot2)
library(haven)
library(readr)
library(texreg)
library(nnet)
```

#Univariate distributions
```{r age, fig.cap="Age Proportion"}
ggplot(cleaned_data2016, aes(x=age, y=..prop.., group=1))+
  geom_bar()+
  theme_bw()
```

#Bivariate plots
```{r vote_age_2016, echo=FALSE, fig.cap="Proportion of voting by age", fig.align="center" }
ggplot(cleaned_data2016, aes(x=vote, y=..prop.., group=1))+
  geom_bar()+
  facet_wrap(~age)+
  coord_flip()+
  labs(x="Candidate choce", y="proportion")
  theme_bw()
```

```{r vote_age_2012, echo-FALSE, fig.cap="Proportion of voting by age",fig.align="center"}  
  ggplot(cleaned_data2012, aes(x=vote, y=..prop.., group=1))+
  geom_bar()+
  facet_wrap(~age)+
  coord_flip()+
  labs(x="Candidate choce", y="proportion")
  theme_bw()
```

Starting at 18 through around the 50 and up age group, Clinton wins more votes, that changes around 50 with the highest proportion of Trump voters being in the 70-75 age range. 
Need to collapse age ranges.

```{r media_age_2016}
cleaned_data2016$media_att_score<- as.numeric(cleaned_data2016$media_att)
model<-lm(media_att, data=cleaned_data2016)

ggplot(cleaned_data2016, aes(x=media_att, y=..prop.., group=1))+
  geom_bar()+
  facet_wrap(~age)+
  coord_flip()+
  labs(x="Media attention", y="proportion")+
  theme_bw()
```
  
```{r media_age_2012}
ggplot(cleaned_data2012, aes(x=media_intINT, y=..prop.., group=1))+
  geom_bar()+
  facet_wrap(~age)+
  coord_flip()+
  labs(x="Media attention Internet", y="proportion")+
  theme_bw()
```


```{r media_party_2012}
ggplot(cleaned_data2012, aes(x=media_intINT, y=..prop.., group=1))+
  geom_bar()+
  facet_wrap(~party)+
  labs(x="Media attention Internet", y="proportion")+
  coord_flip()+
  theme_bw()
```

```{r media_party_2016}
ggplot(cleaned_data2016, aes(x=media_att, y=..prop.., group=1))+
  geom_bar()+
  facet_wrap(~partyreg)+
  coord_flip()+
  labs(x="Media attention by party", y="Proportion")+
  theme_bw()
```


```{r media_vote_2012}
ggplot(cleaned_data2012, aes(x=media_intINT, y=..prop.., group=1))+
  geom_bar()+
  facet_wrap(~vote)+
  coord_flip()+
  labs(x="Media attention Internet", y="proportion")+
  theme_bw()
```

```{r media_vote_2016}
ggplot(cleaned_data2016, aes(x=media_att, y=..prop.., group=1))+
  geom_bar()+
  facet_wrap(~vote)+
  coord_flip()+
  labs(x="Media attention by vote", y="proportion")+
  theme_bw()
```

```{r social_party_2016}
ggplot(cleaned_data2016, aes(x=socialmedia, y=..prop.., group=1))+
  geom_bar()+
  facet_wrap(~vote)+
  coord_flip()+
  labs(x="Social media usage", y="proportion")+
  theme_bw()
```

```{r socialmedia_party}
ggplot(cleaned_data2016, aes(x=socialmedia, y=..prop.., group=1))+
  geom_bar()+
  facet_wrap(~partyreg)+
  labs(x="Social media usage by party", y="proportion")+
  theme_bw()

```

```{r model}
#create continuous age variable

model.full<-multinom(vote~media_att+I(agequant-40)+I((agequant-40)^2)+gender, data=cleaned_data2016)
round(coef(model.full),5)

age <- 18:80
fakedata <- data.frame(media_att=rep(c("None at all","A little","A moderate amount","A lot","A great deal"), length(age)),
                       agequant=rep(age,each=5), gender=rep("Female",5*length(age)))
predicted.probs <- predict(model.full, fakedata, type="probs")

fake.results <- cbind(rbind(fakedata, fakedata, fakedata, fakedata),
                      as.data.frame.table(predicted.probs))
colnames(fake.results) <- c("media_att","age","gender","id","candidate","proportion")

ggplot(fake.results, aes(x=age, y=proportion, color=candidate, group=candidate))+
  geom_line()+
  facet_wrap(~media_att)+
  theme_bw()

ggplot(fake.results, aes(x=age, y=proportion, color=media_att, group=media_att))+
  geom_line()+
  facet_wrap(~candidate)+
  theme_bw()


rm(model.full)
```
```{r model_2012}

#Still giving me 2016 results for some reason 
modelx.full_2012<-multinom(vote~media_intINT+I(agequant-40)+I((agequant-40)^2)+gender, data=cleaned_data2012)
round(coef(modelx.full_2012),5)

age <- 18:80
fakedata_12 <- data.frame(media_intINT=rep(c("None at all","A little","A moderate amount","A lot","A great deal"), length(age)),
                       agequant=rep(age,each=5), gender=rep("Female",5*length(age)))
predicted.probs_12 <- predict(modelx.full_2012, fakedata_12, type="probs")

fake.results_12 <- cbind(rbind(fakedata_12, fakedata_12, fakedata_12, fakedata_12),
                      as.data.frame.table(predicted.probs_12))
colnames(fake.results_12) <- c("media_intINT","age","gender","id","candidate","proportion")


ggplot(fake.results_12, aes(x=age, y=proportion, color=candidate, group=candidate))+
  geom_line()+
  facet_wrap(~media_intINT)+
  theme_bw()

ggplot(fake.results_12, aes(x=age, y=proportion, color=media_intINT, group=media_intINT))+
  geom_line()+
  facet_wrap(~candidate)+
  theme_bw()


```

```{r model_attempt}

model.full<-multinom(media_att~partyreg+vote+age, data=cleaned_data2016)
coef(model.full)

model<-lm(media_att~partyreg+vote+age, data=cleaned_data2016)

```


# Introduction
<a href="#top">Back to top</a>
#takes you back to the top
Use this R Markdown to perform the main analysis for the project. I use this basically as a lab notebook. It contains the main analysis and a variety of sensitivity analysis. The code in this documents serves as a baseline for the eventual tables and figures that will go into the paper. At the same time it will serve as a record of all supplementary analyses performed. 